Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-13 kappa.tex
Add results.

@@

\small{\begin{enumerate}
  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
simulations and wrote the manuscript; C.~Khroulev implemented necessary code
changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
some of the forcing data. All authors contributed to improve the text.
  \item[II] I analysed the weather data, wrote the positive degree day model,
ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
reviews.
  \item[III] I analysed the weather data; I.~Rogozhina and I wrote the manuscript.
  \item[IV] I implemented necessary code changes in PISM, ran the simulations
and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
contributed to improve the text.
\end{enumerate}}

@@

% ----------------------------------------------------------------------
\subsection{Constant-climate runs}
% ----------------------------------------------------------------------

\begin{figure*}
  \centering
  \vspace{-16mm}
  \makebox[0pt]{
  \begin{minipage}[t]{112.5mm}\vspace{0pt}
    \includegraphics{plot-lgm}
  \end{minipage}
  \hspace{1cm}
  \begin{minipage}[t]{47.5mm}\vspace{0pt}
    \caption{Constant-climate simulation of the LGM Cordilleran ice sheet using
             an horizontal resolution of 3\,km. Modelled basal topography,
             cold-based areas (hatches), surface topography (100\,m contours)
             and surface velocities. The simulation is forced by atmospheric
             fields from the NARR \citep{Mesinger.etal.2006} and a temperature
             depression of 7\,\degC, using the set-up described in \CCLI. There
             are several issues with this run. First, all east-facing margins
             are too extensive as compared to available reconstructions.
             Second, marine advance on the Pacific Shelf is too restrictive,
             most notably in the south-western corner where the Puget and Juan
             de Fuca lobes are missing. Third, en-glacial temperatures are
             anomalously low due to brutal application of cold climate
             condition to ice-free bedrock at the inception of the run.
             \todo{Add black frame, and maybe the LGM outline from Dyke?}}
  \end{minipage}
  }
  \vspace{-16mm}
  \label{fig:plot-lgm}
\end{figure*}

@@

@@

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-sdeffect}}
  \caption{Effect of temperature variability on Cordilleran ice sheet
           simulations. Modelled basal topography and surface topography
           (250\,m contours). The simulations use atmospheric forcing from the
           NARR \citep{Mesinger.etal.2006}, a temperature depression of
           7\,\degC, and different treatments of temperature variability
           including
           \textbf{(a)} no daily variability, $\sigma=0$;
           \textbf{(b)} a constant, mean summer (JJA), model domain-averaged 
           standard deviation value, $\sigma=3.09$\,K;
           \textbf{(c)} monthly standard deviation grids derived from the NARR
           and
           \textbf{(d)} monthly grids for ice-free terrain and the
           Greenland ice sheet, ERA-40-based \citep{Uppala.etal.2005}
           parametrisation in \PSDP, ${\sigma=-0.15\cdot{T_m}+1.66}$.
           Differences in surface elevation are displayed relative to case (c).
           Note the different colour scales.}
  \label{fig:plot-sdeffect}
\end{figure}

% ----------------------------------------------------------------------
\subsection{Transient-climate runs}
% ----------------------------------------------------------------------

@@

% References
\newcommand{\urlprefix}[0]{}  % remove default "URL" prefix
\bibliographystyle{copernicus}
\bibliography{refs/references.bib}

>>>>

% Github links
\newcommand{\issue}[1]{\href{https://github.com/pism/pism/issues/#1}{\##1}}

@@

\small{\begin{enumerate}
  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
simulations and wrote the manuscript; C.~Khroulev implemented necessary code
changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
some of the forcing data. All authors contributed to improve the text.
  \item[II] I analysed the weather data, wrote the positive degree day model,
ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
reviews.
  \item[III] I.~Rogozhina provided the original idea; I analysed the weather
data; and we wrote the manuscript together.
  \item[IV] I implemented necessary code changes in PISM, ran the simulations
and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
contributed to improve the text.
\end{enumerate}}

@@

% ----------------------------------------------------------------------
\subsection{Constant-climate simulations}
% ----------------------------------------------------------------------

\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}
\bigskip

As a first step towards simulations of the Cordilleran ice sheet through the
last glacial cycle, I have chosen to run the ice sheet model under a constant
climate, targeting the LGM as the period for which ice sheet limits are best
known. These simulations, presented in \CCLI, use constant temperature offsets
from the present climate state. They start from ice-free conditions and run for
10\,ka, a period chosen as representative of the ice-sheet build-up from
assumed, negligible ice cover. Although this assumption is disproved in \CCYC,
the simplistic set-up used in \CCLI\ allowed to test model sensitivity to the
choice of present-day climate forcing data, without involving too many
additional parameters.

Discrepancies between datasets used (\cref{tab:datasources}) cause much
variability in modelled ice sheet geometries (\CCLI, Figs.~5--7). In
particular, continental parts of the ice-sheet are strongly dependent on input
precipitation patterns, which are often poorly resolved by atmospheric
reanalysis fields used to force the model (\CCLI, Figs.~10--14). The closest
match between model output and the reconstructed ice margin \citep{Dyke.2004}
was obtained using atmospheric forcing from the North American Regional
Reanalysis \citep[NARR,][]{Mesinger.etal.2006} and a temperature depression of
7\,\degC\ (\CCLI, Fig.~15). The relevant simulation is reproduced here, using a
higher horizontal resolution of 3\,km (\cref{fig:plot-lgm}). The NARR is the
dataset that I have retained to force simulations of the Cordilleran ice sheet
presented here (e.g., \cref{fig:plot-lgm}) and in \CCYC.

\begin{figure*}
  \centering
  \vspace{-16mm}
  \makebox[0pt]{
  \begin{minipage}[t]{112.5mm}\vspace{0pt}
    \includegraphics{plot-lgm}
  \end{minipage}
  \hspace{1cm}
  \begin{minipage}[t]{47.5mm}\vspace{0pt}
    \caption{Constant-climate simulation of the LGM Cordilleran ice sheet using
             an horizontal resolution of 3\,km. Modelled basal topography,
             cold-based areas (hatches), surface topography (100\,m contours)
             and surface velocities. The simulation is forced by atmospheric
             fields from the NARR \citep{Mesinger.etal.2006} and a temperature
             depression of 7\,\degC, using the set-up described in \CCLI. There
             are several issues with this run. First, all east-facing margins
             are too extensive as compared to available reconstructions.
             Second, marine advance on the Pacific Shelf is too restrictive,
             most notably in the south-western corner where the Puget and Juan
             de Fuca lobes are missing. Third, en-glacial temperatures are
             anomalously low due to brutal application of cold climate
             condition to ice-free bedrock at the inception of the run.
             \todo{Add black frame, and maybe the LGM outline from Dyke?}}
    \label{fig:plot-lgm}
  \end{minipage}
  }
  \vspace{-16mm}
\end{figure*}

@@

\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}
\bigskip

\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}
\bigskip

Next to the atmospheric forcing, another critical source of uncertainty in
palaeo-ice sheet modelling is the choice of parameters for the surface mass
balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
are the degree-day factors for snow, $F_s$, and ice, $F_i$, and the standard
deviation of near-surface air temperature, $\sPDD$. Although the first two can
only be deduced from laborious surface mass-balance measurements on present-day
glaciers \citep[e.g.,][]{Shea.etal.2009}, $\sPDD$ can be estimated from
observed, modelled, or reanalysed temperature time-series
\citep[e.g.,][]{Fausto.etal.2011}.

Using temperature time-series from the ERA-Interim reanalysis
\citep{Dee.etal.2011}, this is what is done of \PSDV, where I show, that,
although often assumed constant, $\sPDD$ greatly varies through the year and
around the globe, most particularly in the polar regions
(\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using a simple surface mass balance
model, I then demonstrate the effect of these variations on the computation of
positive degree-days (\PSDV, Fig.~2), and surface mass-balance (\PSDV, Fig.~3).

The analysis is taken further in \PSDP, using re-analysed air temperature from
ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial resolution, the
ERA-40 time series have a longer time span than that of the ERA-Interim, which
is necessary to compute standard deviation values that are representative of
thelong-term daily variability \citep{Rogozhina.Rau.2014}. In \PSDP, we
demonstrated that, over the Greenland ice sheet, there exist an inverse
correlation between $\sPDD$ values and the monthly temperature means, $T_m$
(\PSDP, Fig.~1). This correlation could potentially be used as a
parametrisation, which can be expressed as (\PSDP, Eq.~5)

\begin{equation}
    \label{eqn:sigma}
    \sPDD = -0.15 \cdot T_{m} + 1.66\,.
\end{equation}

The effect of the PDD standard deviation, $\sPDD$, on simulations of the
Cordilleran ice sheet is not shown in \PSDV\ and \PSDP. However, following
these two studies, a natural step further was to implement spatially and
seasonally variable $\sPDD$ (PISM issue \issue{179}), and the parametrisation
expressed by \cref{eqn:sigma} over ice-filled grid cells (\issue{265}), in
PISM. The effect of various temperature variability modelling approaches, and
associated errors, is demonstrated here using a constant-climate set-up similar
to that of \CCLI\ (\cref{fig:plot-sdeffect}).

@@

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-sdeffect}}
  \caption{Effect of temperature variability on Cordilleran ice sheet
           simulations. Modelled basal topography and surface topography
           (250\,m contours). The simulations use atmospheric forcing from the
           NARR \citep{Mesinger.etal.2006}, a temperature depression of
           7\,\degC, and different treatments of temperature variability
           including
           \textbf{(a)} no daily variability, $\sPDD=0$;
           \textbf{(b)} a constant, mean summer (JJA), model domain-averaged
           standard deviation value, $\sPDD=3.09$\,K;
           \textbf{(c)} monthly standard deviation grids derived from the NARR
           and
           \textbf{(d)} monthly grids for ice-free terrain and the
           Greenland ice sheet, ERA-40-based \citep{Uppala.etal.2005}
           parametrisation in \PSDP, ${\sPDD=-0.15\cdot{T_m}+1.66}$.
           Differences in surface elevation are displayed relative to case (c).
           Note the different colour scales.}
  \label{fig:plot-sdeffect}
\end{figure}

% ----------------------------------------------------------------------
\subsection{Glacial cycle simulations}
% ----------------------------------------------------------------------

\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}
\bigskip

From the start, the primary goal of this thesis has been to perform simulations of the Cordilleran ice sheet through the last glacial cycle. This is the topic of \CCYC, where monthly air temperature, precipitation, and PDD standard deviation computed from the NARR are amended by palaeo-temperature time-series from six proxy records to produce a time-dependent climate forcing (\CCYC, Fig.~2). The resulting simulations all produce two major glaciations between MISs~4 and 2. However the differences in timing and magnitude of these events is considerable (\CCYC, Fig.~3--4).

Geological evidence regarding the timing of glaciation allowed us to select two
simulations, driven by the GRIP \citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice-core records, as producing a more realistic reconstruction of the last glacial cycle. I have re-run these two simulations at a higher resolution of 6\,km, to allow for a more detailed study of the primary locations of transient ice cap cover, stability of major ice-dispersal centres, and patterns of deglaciation (\CCYC, Fig.~5--15). Modelled ice volume evolution and snapshots of ice flow patterns from one of these two simulations are reproduced here (\cref{fig:plot-snapshots}).

\todo{Verify that figure numbers remain correct as \CCYC\ evolves.}

@@

% References
\renewcommand{\urlprefix}[0]{}  % remove default "URL" prefix
\bibliographystyle{copernicus}
\bibliography{refs/references.bib}
